---
title: "Test generative model using CmdStan"
author: "Jacqueline Buros"
date: "May 20, 2016"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r init, results = 'hide', echo = F}
suppressMessages(suppressWarnings({
  rm(list=ls())
  library(dplyr)
  library(ggplot2)
  library(rstan)
  library(survival)
  library(arm)
  library(lme4)
  rstan_options(auto_write = TRUE)
  options(mc.cores = min(parallel::detectCores(),3))
  source('simulate-data.function.R')
  source('prep-data.function.R')
  source('make-data-plots.function.R')
}))
```

## Intro

This is a short analysis to test various options for estimating the "generative" model in Stan,
and specifically using CmdStan. 

There are several reasons why I'm looking at using CmdStan:

1. Rstan is prone to fail/crash at some point, particularly for long-running jobs
    - preliminary testing shows that these models will take a while to run
2. Better diagnostics & more up to date codebase (e.g. support for stiff odes) with CmdStan
3. Having problems running R on dev nodes (more about this later ... )

## Simulated data

Data were simulated according to the `simulate_data()` function. See that function for more details.

Here we will use standard functions designed to help in post-processing of the simulated data.

```{r sim-data}
set.seed(1234)
d <- simulate_data(n = 100, max_size = 4000, max_t = 50, failure_threshold = 4, progression_threshold = 3)
make_data_plots(d)
```

We will next prepare two data files for analysis - one _per patient_, which will be suitable for survival analysis, and another _per observation_. This latter, _per-observation_ dataset is the one we will predominantly pass to cmdstan.

```{r prep-data}
res <- prep_data(d)
adata <- res$per_observation
survd <- res$per_patient
rm(res)
```

## Standard models 

Just for kicks, let's run a few standard models on this data. 

We will focus on two covariates in particular - 

1. Initial tumor size (observed)
2. Avg rate of change in tumor size: (final - initial) / time

We can look at these in terms of failure (survival), or the combintion of disease progression and/or failure. 

#### with failure as the outcome 

Looking at initial tumor size as a predictor : 

```{r surv1}

survfit1 <- coxph(
  formula = Surv(first_failure, failure_status) ~ rescaled_init_size
  , data = survd
  )
print(survfit1)
```


And, including change in tumor size : 

```{r surv2}

survfit <- coxph(
  formula = Surv(first_failure, failure_status) ~ rescaled_init_size + rescaled_growth_rate
  , data = survd
  )
print(survfit)
```

#### with progression/failure as the outcome 

Looking at initial tumor size as a predictor : 

```{r surv3}

survfit <- coxph(
  formula = Surv(first_failure_or_progression, failure_or_progression_status) ~ rescaled_init_size
  , data = survd
  )
print(survfit)
```


And, including change in tumor size : 

```{r surv4}
survfit <- coxph(
  formula = Surv(first_failure_or_progression, failure_or_progression_status) ~ rescaled_init_size + rescaled_growth_rate
  , data = survd
  )
print(survfit)
```


## Stan code for basic survival model 

We have Stan code for the basic survival model. Let's run it now just to get a sense of how it works.

```{r surv-model}
standata <- list(
  N = nrow(adata)
  , S = max(adata$patid)
  , T = max(adata$t)
  , X = 1
  , s = adata$patid
  , t = adata$t
  , event = adata$failure
  , covars = adata %>% dplyr::select(rescaled_init_size)
)

testfit <- stan('long_surv.stan', data = standata, chains = 1, iter = 10)

stanfit <- stan('long_surv.stan', data = standata, chains = 3, iter = 1000)

print(stanfit, 'beta')
```

Now, this is not _identical_ to the survival analysis, but we should see very similar results to the estimates we got from `coxph`.

Let's confirm this now.

```{r review-estimates}
print(survfit1)
```

## Semi-competing risks model 

We additionally have Stan code for a semi-competing risks model. This analysis will be summarized separately.

## Generative model 

We are finally working on what I'll call the "generative model" -- this is the model that more closely mimics the data generating process we hypothesized might be at work.

The code right now is making use of the `integrate_ode` function in Stan. 

We will first review how this works & then test out the implementation using CmdStan.

### Ex 1: using stan to simulate data

Here is the stan file :

```{r echo-stan-sim}
file_path <- 'generative_model_sim_data.stan'
lines <- readLines(file_path, encoding="ASCII")
for (n in 1:length(lines)) cat(lines[n],'\n')
```

```{r sim-data-stan}
## pick a random patient; pass these parameters to the stan file 
sample_data <- adata %>% semi_join(adata %>% sample_n(1) %>% dplyr::select(patid), by = 'patid')

sample_params <- list(
  N_obs = nrow(sample_data)
  , obs_t = sample_data$t
  , init_vol = unique(sample_data$init_size)
  , growth_rate = unique(sample_data$growth_rate)
  , max_size = 4000
)
testfit <- stan('generative_model_sim_data.stan', data = sample_params, chains = 1, iter = 100, algorithm = 'Fixed_param')
print(testfit)
```

### Ex 2: estimating parameters for a single patient 

Next we want to try to use `Stan` to estimate the growth rate for a single patient's tumor.

We will try to set this up as a "best case" scenario, where we will tell Stan what the max-size is, and we will also estimate this using actual _tumor size_ measurements, instead of the diameter measurements (with measurement error) that were simulated. 

Here is the stan file we'll be using to estimate this:

```{r echo-stan-single-pat}
file_path <- 'generative_model_single_obs_more_params.stan'
lines <- readLines(file_path, encoding="ASCII")
for (n in 1:length(lines)) cat(lines[n],'\n')
```

#### invoking CmdStan

The process for invoking this from CmdStan involves a few steps : 

1. Write out data to file
2. Call "make" to translate *.stan -> *.cpp -> executable
3. invoke executable on the data file, with params (possibly multiple chains)

Written in R code, this looks like : 

```{r invoke-cmdstan, echo=TRUE, warning = FALSE, eval = TRUE}
## prep data for stan
growthdata <- list(
  N_obs = nrow(sample_data)
  , obs_t = sample_data$t
  , obs_size = sample_data$tumor_size
  , max_size = 4000
)
## call using cmdstan instead of rstan. more reliable for longer-running models
stan_home <- '/usr/local/Cellar/cmdstan/2.9.0'
modelpath <- file.path(getwd(),'generative_model_single_obs_more_params.stan')
modelpath <- gsub(modelpath, pattern = "(.*)\\.stan", replacement = '\\1')
modelname <- gsub(modelpath, pattern = ".*\\/([^\\/.]+)$", replacement = "\\1")
datafile <- file.path(getwd(),paste0(modelname,'.data.R'))
## write data to disk
with(growthdata, stan_rdump(names(growthdata), file = datafile))
## translate stan -> c++; compile to executable
system(paste0('(cd ',stan_home,' && make ',modelpath,')'))
## test on 10 iterations

system(paste0('./',modelname,' diagnose data file=',datafile))
system(paste0('./',modelname,' sample num_samples=5 num_warmup=5 random seed=12345 data file=',datafile))
```

#### invoking rstan

```{r invoke-rstan, echo = TRUE, eval = FALSE}
growthdata <- list(
  N_obs = nrow(sample_data)
  , obs_t = sample_data$t
  , obs_size = sample_data$tumor_size
  , max_size = 4000
)
testfit <- stan('generative_model_single_obs_more_params.stan', data = growthdata, iter=10, chains = 1)
```


